Visual Question Answering with DeepProbLog

### 2022-03-21 paper/paper.tex
Review whole paper again

\section{ORGANIZATION OF THE PAPER}
This paper will first provide some necessary background in Section \ref{sec:literature_survey} on certain types of datasets that are typically used for Visual Question Answering purposes, along with introducing some results of other researchers. Then, Section \ref{sec:approach} will dive deeper in the different components constituting the overall system. Section \ref{sec:experiments}, furthermore, provides some results of conducted experiments. A main focus in this paper is to outline the advantages of using a Neuro-Symbolic AI approach (offered by the DeepProbLog framework), instead of a purely neural network based approach. These experiments are listed in Section \ref{sec:experiments}. Finally, the main findings of this paper are repeated in Section \ref{sec:conclusions}.

\section{LITERATURE SURVEY}
\label{sec:literature_survey}
The application focuses on Visual Question Answering (VQA), for which huge datasets are present, along with very sophisticated methods. The best known dataset for VQA is CLEVR \cite{clevr_dataset}, which contains 100k images with one million questions. An example image is given in Figure \ref{fig:sample_image_clevr}, while example questions are:
\begin{itemize}
    \item Are there an equal number of large things and metal spheres?
    \item What size is the cylinder that is left of the brown metal thing that is left of the big sphere?
    \item How many objects are either small cylinders or metal things?
\end{itemize}

@@

This paper is inspired on the CLEVR dataset, but uses however a much more simplified version. In essence, it is almost like the Sort-Of-CLEVR dataset \cite{sort_of_clevr_dataset}. This Sort-Of-CLEVR dataset contains images such as in Figure \ref{fig:sample_image_sort_of_clevr}, while asking questions such as:
\begin{itemize}
    \item Non-relational questions: the shape, horizontal or vertical location of an object.
    \item Relational questions: shape of the closest/furthest object to the object under investigation, or the number of objects with the same shape.
\end{itemize}

@@

A. Santoro et al. conducted similar experiment with the Sort-of-CLEVR dataset as in this paper, where they were able to achieve an accuracy of 63\% on relational questions with CNN's. In contrast, an accuracy of 94\% for both relational and non-relational questions was achieved with CNN's complemented with an RN. For them, augmenting the model with a relational module, such as an RN, turned out to be sufficient to overcome the hurdle of solving relational questions \cite{sort_of_clevr_dataset}.

@@

\subsection{Generation of data}
As mentioned in Section \ref{sec:literature_survey}, the data used in this application is based on the Sort-Of-CLEVR dataset, with one extra simplification. Given that the logical part will have to decide whether an object is for example located on the left side of an image, the neural network will have to convey positional information to the logical part. Hence, each discrete position will have to be encoded by a possible outcome of the neural network. Therefore, objects may only be located at certain places in a grid. In this paper, results on a grid of 2x2 and 6x6 are discussed.

@@

These images are accompanied with a question about a random object, which can be one of the following:
\begin{itemize}
    \item Non-relational - What is the shape of this object \footnote{The shape will be either a square or a circle.}?
    \item Non-relational - Is this object located on the left side of the image?
    \item Non-relational - Is this object located on the bottom side of the image?
    \item Relational - How many objects have the same shape as this object?
\end{itemize}

@@

\section{EXPERIMENTS}
\label{sec:experiments}
The main focus of this paper is to outline the advantages of using a Neuro-Symbolic AI approach (offered by the DeepProbLog framework), instead of a purely neural network based approach. Therefore, another purely neural network based approach had to be implemented. Not all details will be listed here, but the main idea is that the image has to be fused with the question, after which a prediction can be made\footnote{Possible answers on the questions are: rectangle, circle, yes, no, 1, 2, 3, 4, 5 and 6. Hence, the neural network will have 10 output nodes.}. The general structure of this network is given in Figure \ref{fig:pure_nn_network}.

@@

\begin{figure}[htp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{loss_curve_pure_NN_2x2.png} 
    \captionsetup{justification=centering}
    \caption{Pure NN based approach 2x2: Loss curve}
    \label{fig:loss_curve_pure_NN_2x2}
    \end{center}
\end{figure}
An extremely important remark to be made is the difference between `Number of iterations' and `Number of epochs'. By the number of iterations, the number of forward and backward passes of a batch (with size 32) is meant, whereas the number of epochs denote the number of times all the images of the training set are forward and backwardly passed. In this application, a training size of 10 000 was used, hence one epoch consists of 312.5 iterations.

@@

On the other hand, one has to consider the actual running times for those iterations. DeepProbLog takes around 10 minutes to finish its 160 iterations, while the purely neural network based approach requires around 5 minutes to finish 7 800 iterations. Taking into consideration that the purely neural network based approach can be accelerated massively (by using GPU's), while DeepProbLog can't\footnote{DeepProbLog offers the ability to send the CNN to a GPU for faster inference, however, the arithmetic operators (i.e. semirings) of DeepProbLog work on the CPU. These arithmetic operators possess by far the highest computational cost.}, it is clear that DeepProbLog trains much more targeted, but is computationally extremely heavy (at least for now).

@@

In these experiments, the training time of the purely neural network approach was 20 minutes\footnote{For fair speed estimation, training has been conducted on the CPU.}, while the training time of the DeepProbLog approach was a little under eight hours\footnote{However, it should be mentioned that approximately half of the time was spent on calculating the accuracy each time, since the whole testing testing set had to be forwarded through the network, while for the training iterations this is only one batch.}.

@@

Another observation is that DeepProbLog is much better in answering the questions `What shape has the object under investigation?' and `Is the object located on the left hand side (or on the bottom side) of the image?' then a purely neural network approach is. This makes sense, since DeepProbLog can use its knowledge base to derive these properties, once the position (and shape) of the given object is determined. The observation that a pure neural network based approach has a much harder time to distinguish these cases, was also observed by \cite{sort_of_clevr_dataset}, where they achieved an accuracy of 52.3\% on these questions with a pure neural network based approach.

@@

\section*{APPENDIX}
\subsection{Logical rule encodings}
\label{appendix:logical_rule_encodings}
\inputminted[breaklines]{prolog}{"/home/jorrit/Data/KU Leuven/Semester 12/Capita Selecta H05N0a/deepproblog/src/deepproblog/examples/SORTOFCLEVR/model.pl"}

>>>>

\section{ORGANIZATION OF THE PAPER}
This paper will first provide some necessary background in Section \ref{sec:literature_survey} on certain types of datasets that are typically used for Visual Question Answering purposes, along with introducing some results of other researchers. Then, Section \ref{sec:approach} will dive deeper in the different components constituting the overall system. A main focus of this paper is to outline the advantages of using a Neuro-Symbolic AI approach (offered by the DeepProbLog framework), compared to a purely neural network based approach. These experiments are listed in Section \ref{sec:experiments}. Finally, the main findings of this paper are repeated in Section \ref{sec:conclusions}.

\section{LITERATURE SURVEY}
\label{sec:literature_survey}
The application focuses on Visual Question Answering (VQA), for which huge datasets are present, along with very sophisticated methods. The best known dataset for VQA is CLEVR \cite{clevr_dataset}, which contains 100k images accompanied by one million questions. An example image is given in Figure \ref{fig:sample_image_clevr}, while example questions are:
\begin{itemize}
    \item Are there an equal number of large things and metal spheres?
    \item What size is the cylinder that is left of the brown metal thing that is left of the big sphere?
    \item How many objects are either small cylinders or metal things?
\end{itemize}

@@

This paper is inspired on the CLEVR dataset, but uses however a much more simplified version. In essence, it is almost like the Sort-Of-CLEVR dataset \cite{sort_of_clevr_dataset}. This Sort-Of-CLEVR dataset contains images as in Figure \ref{fig:sample_image_sort_of_clevr}, while asking questions such as:
\begin{itemize}
    \item Non-relational questions: the shape, horizontal or vertical location of an object.
    \item Relational questions: shape of the closest/furthest object to the object under investigation, or the number of objects with the same shape.
\end{itemize}

@@

A. Santoro et al. conducted similar experiments with the Sort-of-CLEVR dataset as in this paper, where they were able to achieve an accuracy of 63\% on relational questions with CNN's. In contrast, an accuracy of 94\% for both relational and non-relational questions was achieved with CNN's, complemented with an RN. For them, augmenting the model with a relational module, such as an RN, turned out to be sufficient to overcome the hurdle of solving relational questions \cite{sort_of_clevr_dataset}. They were the only researchers who undertook experiments on a dataset resembling the one from this paper, and hence are the only reference point.

@@

\subsection{Generation of data}
As mentioned in Section \ref{sec:literature_survey}, the data used in this application is based on the Sort-Of-CLEVR dataset, with one extra simplification. Given that the logical part will have to decide whether an object is for example located on the left side of an image, the neural network will have to convey positional information to the logical part. Hence, each discrete position will have to be encoded by a possible outcome of the neural network. Therefore, objects may only be located at certain positions in a grid. In this paper, results on a grid of 2x2 and 6x6 are discussed.

@@

These images are accompanied with a question about a random object, which can be one of the following:
\begin{itemize}
    \item Non-relational - What is the shape of this object?\footnote{The shape will be either a square or a circle.}
    \item Non-relational - Is this object located on the left side of the image?
    \item Non-relational - Is this object located on the bottom side of the image?
    \item Relational - How many objects have the same shape as this object?
\end{itemize}

@@

\section{EXPERIMENTS}
\label{sec:experiments}
The main focus of this paper is to outline the advantages of using a Neuro Symbolic AI approach (offered by the DeepProbLog framework), instead of a purely neural network based approach. Therefore, a neural network based approach had to be implemented. Not all details will be listed here, but the main idea is that the image has to be fused with the question, after which a prediction can be made\footnote{Possible answers on the questions are: rectangle, circle, yes, no, 1, 2, 3, 4, 5 and 6. Hence, the neural network will have 10 output nodes.}. The general structure of this network is given in Figure \ref{fig:pure_nn_network}.

@@

\begin{figure}[htp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{loss_curve_pure_NN_2x2.png} 
    \captionsetup{justification=centering}
    \caption{Pure NN based approach 2x2: Loss curve}
    \label{fig:loss_curve_pure_NN_2x2}
    \end{center}
\end{figure}
An extremely important remark to be made is the difference between `number of iterations' and `number of epochs'. By the number of iterations, the number of forward and backward passes of a batch (with size 32) is meant, whereas the number of epochs denotes the number of times all the images of the training set are forward and backwardly passed. In this application, a training size of 10 000 was used, hence one epoch consists of 312.5 iterations.

@@

On the other hand, one has to consider the actual running times for those iterations. DeepProbLog takes around 10 minutes to finish its 160 iterations, while the purely neural network based approach only requires around 5 minutes to finish 7 800 iterations. Taking into consideration that the purely neural network based approach can be accelerated massively (by using GPU's), while DeepProbLog can't\footnote{DeepProbLog offers the ability to send the CNN to a GPU for faster inference, however, the arithmetic operators (i.e. semirings) of DeepProbLog work on the CPU. These arithmetic operators possess by far the highest computational cost.}, it is clear that DeepProbLog trains much more targeted, but is computationally extremely heavy (at least for now).

@@

In these experiments, the training time of the purely neural network approach was 20 minutes\footnote{For fair speed estimation, training has been conducted on the CPU.}, while the training time of the DeepProbLog approach was a little under eight hours\footnote{However, it should be mentioned that approximately half of the time was spent on calculating the accuracy each time, since the whole testing set had to be forwarded through the network, while for the training iterations this is only one batch.}.

@@

Another observation is that DeepProbLog is much better in answering the questions `What shape has the object under investigation?' and `Is the object located on the left hand side (or on the bottom side) of the image?' then a purely neural network approach is. This makes sense, since DeepProbLog can use its knowledge base to derive these properties, once the position (and shape) of the given object is determined. The observation that a pure neural network based approach has a much harder time to distinguish these cases, was also observed by A. Santoro et al. \cite{sort_of_clevr_dataset}, where they achieved an accuracy of 52.3\% on these questions with a pure neural network based approach. DeepProbLog was not able to achieve the accuracy of 94\% that these researchers achieved, which may be due to the inherent training cost of the algebraic operators, as well as due to less resources and less hyperparameter tuning, among many other unknown variables.

@@

\section*{APPENDIX}
\subsection{Logical rule encodings}
Note: it has been tried to use cuts (with the `cut' library), given that the neural network is seemingly invoked again if the first of a dual predicate fails. However, this turned out to not yield any performance improvement and has been abandoned.
\label{appendix:logical_rule_encodings}
\inputminted[breaklines]{prolog}{"/home/jorrit/Data/KU Leuven/Semester 12/Capita Selecta H05N0a/deepproblog/src/deepproblog/examples/SORTOFCLEVR/model.pl"}
